---
title: "Modeling Deaths"
author: "Safiya Sirota"
date: "2023-03-30"
output: html_document
---
# Set up

## Load in Anja's and Ting-Hsuan's code

```{r}
source("anja.R")
source("tinghsuan_wave1case.R")
```

# Wave 1

## Overwrite `df_wave` to consider death counts instead of death counts

Changing to deaths and setting up the data for each borough in wave 1

```{r}
# All of NYC
df_wave1_nyc <- df %>%
  filter(date >= start_date & date <= end_date) %>%
  select(date, death_count) %>%
  mutate(
    t_i = get_diff(start_date, date),
    Y_i = cumsum(death_count)
  ) %>%
  select(!death_count)

# Bronx
df_wave1_bx <- df %>%
  filter(date >= start_date & date <= end_date) %>%
  select(c(date, bx_death_count)) %>%
  mutate(
    t_i = get_diff(start_date, date),
    Y_i = cumsum(bx_death_count)
  ) %>%
  select(!bx_death_count)

# Brooklyn
df_wave1_bk <- df %>%
  filter(date >= start_date & date <= end_date) %>%
  select(c(date, bk_death_count)) %>%
  mutate(
    t_i = get_diff(start_date, date),
    Y_i = cumsum(bk_death_count)
  ) %>%
  select(!bk_death_count)

# Manhattan
df_wave1_mn <- df %>%
  filter(date >= start_date & date <= end_date) %>%
  select(c(date, mn_death_count)) %>%
  mutate(
    t_i = get_diff(start_date, date),
    Y_i = cumsum(mn_death_count)
  ) %>%
  select(!mn_death_count)

# Queens
df_wave1_qn <- df %>%
  filter(date >= start_date & date <= end_date) %>%
  select(c(date, qn_death_count)) %>%
  mutate(
    t_i = get_diff(start_date, date),
    Y_i = cumsum(qn_death_count)
  ) %>%
  select(!qn_death_count)

# Staten Island
df_wave1_si <- df %>%
  filter(date >= start_date & date <= end_date) %>%
  select(c(date, si_death_count)) %>%
  mutate(
    t_i = get_diff(start_date, date),
    Y_i = cumsum(si_death_count)
  ) %>%
  select(!si_death_count)
```

### Check cumulative deaths over wave 1 for all of NYC

```{r}
df_wave1_nyc %>% 
  ggplot(aes(x = date, y = Y_i)) +
  geom_point() +
  scale_x_date(date_breaks = "1 month", date_labels = "%b")
```

### Run N-R

Running N-R for each borough and all of NYC using code from loaded R files

```{r}
# All of NYC
start_nyc <- get_start(df_wave1_nyc, infl_point = "2020-04-10")
res_nyc <- NewtonRaphson(df_wave1_nyc, start_nyc)
final_nyc <- res_nyc[nrow(res_nyc), 4:7]

# Bronx
start_bx <- get_start(df_wave1_bx, infl_point = "2020-04-10")
res_bx <- NewtonRaphson(df_wave1_bx, start_bx)
final_bx <- res_bx[nrow(res_bx), 4:7]

# Brooklyn
start_bk <- get_start(df_wave1_bk, infl_point = "2020-04-10")
res_bk <- NewtonRaphson(df_wave1_bk, start_bk)
final_bk <- res_bk[nrow(res_bk), 4:7]

# Manhattan
start_mn <- get_start(df_wave1_mn, infl_point = "2020-04-10")
res_mn <- NewtonRaphson(df_wave1_mn, start_mn)
final_mn <- res_mn[nrow(res_mn), 4:7]

# Queens
start_qn <- get_start(df_wave1_qn, infl_point = "2020-04-10")
res_qn <- NewtonRaphson(df_wave1_qn, start_qn)
final_qn <- res_qn[nrow(res_qn), 4:7]

# Staten Island
start_si <- get_start(df_wave1_si, infl_point = "2020-04-10")
res_si <- NewtonRaphson(df_wave1_si, start_si)
final_si <- res_si[nrow(res_si), 4:7]
```

### Plot final curves

```{r}
# Bronx
death_wave1_bx <- df_wave1_bx %>%
  mutate(
    Y_i_hat_start = N(t_i, start_bx),
    Y_i_hat_final = N(t_i, final_bx)
  ) %>%
  pivot_longer(Y_i_hat_start:Y_i_hat_final, names_to = "type", values_to = "value") %>%
  ggplot(aes(x = t_i)) + 
  geom_point(aes(y = Y_i), size = 1) +
  geom_line(aes(y = value, color = type)) +
  labs(title = "Bronx", x ="Days (since Feb 29, 2020)", y = "Cumulative deaths") +   
  scale_color_discrete(name = "Fitted model", labels = c("Final", "Start")) 

# Brooklyn
death_wave1_bk <- df_wave1_bk %>%
  mutate(
    Y_i_hat_start = N(t_i, start_bk),
    Y_i_hat_final = N(t_i, final_bk)
  ) %>%
  pivot_longer(Y_i_hat_start:Y_i_hat_final, names_to = "type", values_to = "value") %>%
  ggplot(aes(x = t_i)) + 
  geom_point(aes(y = Y_i), size = 1) +
  geom_line(aes(y = value, color = type)) +
  labs(title = "Brooklyn", x ="Days (since Feb 29, 2020)", y = "Cumulative deaths") +   
  scale_color_discrete(name = "Fitted model", labels = c("Final", "Start")) 

## Manhattan
death_wave1_mn <- df_wave1_mn %>%
  mutate(
    Y_i_hat_start = N(t_i, start_mn),
    Y_i_hat_final = N(t_i, final_mn)
  ) %>%
  pivot_longer(Y_i_hat_start:Y_i_hat_final, names_to = "type", values_to = "value") %>%
  ggplot(aes(x = t_i)) + 
  geom_point(aes(y = Y_i), size = 1) +
  geom_line(aes(y = value, color = type)) +
  labs(title = "Manhattan", x ="Days (since Feb 29, 2020)", y = "Cumulative deaths") +   
  scale_color_discrete(name = "Fitted model", labels = c("Final", "Start")) 

## Queens
death_wave1_qn <- df_wave1_qn %>%
  mutate(
    Y_i_hat_start = N(t_i, start_qn),
    Y_i_hat_final = N(t_i, final_qn)
  ) %>%
  pivot_longer(Y_i_hat_start:Y_i_hat_final, names_to = "type", values_to = "value") %>%
  ggplot(aes(x = t_i)) + 
  geom_point(aes(y = Y_i), size = 1) +
  geom_line(aes(y = value, color = type)) +
  labs(title = "Queens", x ="Days (since Feb 29, 2020)", y = "Cumulative deaths") +   
  scale_color_discrete(name = "Fitted model", labels = c("Final", "Start")) 

## Staten Island
death_wave1_si <- df_wave1_si %>%
  mutate(
    Y_i_hat_start = N(t_i, start_si),
    Y_i_hat_final = N(t_i, final_si)
  ) %>%
  pivot_longer(Y_i_hat_start:Y_i_hat_final, names_to = "type", values_to = "value") %>%
  ggplot(aes(x = t_i)) + 
  geom_point(aes(y = Y_i), size = 1) +
  geom_line(aes(y = value, color = type)) +
  labs(title = "Staten Island", x ="Days (since Feb 29, 2020)", y = "Cumulative deaths") +   
  scale_color_discrete(name = "Fitted model", labels = c("Final", "Start")) 

## NYC
death_wave1_nyc <- df_wave1_nyc %>%
  mutate(
    Y_i_hat_start = N(t_i, start_nyc),
    Y_i_hat_final = N(t_i, final_nyc)
  ) %>%
  pivot_longer(Y_i_hat_start:Y_i_hat_final, names_to = "type", values_to = "value") %>%
  ggplot(aes(x = t_i)) + 
  geom_point(aes(y = Y_i), size = 1) +
  geom_line(aes(y = value, color = type)) +
  labs(title = "New York City", x ="Days (since Feb 29, 2020)", 
       y = "Cumulative deaths") +   
  scale_color_discrete(name = "Fitted model", labels = c("Final", "Start")) 

ggarrange(death_wave1_nyc, death_wave1_bx, death_wave1_bk, death_wave1_mn, death_wave1_qn, death_wave1_si,  nrow = 2, ncol = 3, common.legend = TRUE, legend = "bottom") 
```

# Wave 2

## Reset start and end dates

Loading in wave2death.R to reset start and end dates and to replace the `get_start` function with the proper method/parameters for Wave 2

```{r}
source("wave2case.R")
```

## Set up data

```{r}
# All of NYC
df_wave2_nyc <- df %>%
  filter(date >= start_date & date <= end_date) %>%
  select(date, death_count) %>%
  mutate(
    t_i = get_diff(start_date, date),
    Y_i = cumsum(death_count)
  ) %>%
  select(!death_count)

# Bronx
df_wave2_bx <- df %>%
  filter(date >= start_date & date <= end_date) %>%
  select(c(date, bx_death_count)) %>%
  mutate(
    t_i = get_diff(start_date, date),
    Y_i = cumsum(bx_death_count)
  ) %>%
  select(!bx_death_count)

# Brooklyn
df_wave2_bk <- df %>%
  filter(date >= start_date & date <= end_date) %>%
  select(c(date, bk_death_count)) %>%
  mutate(
    t_i = get_diff(start_date, date),
    Y_i = cumsum(bk_death_count)
  ) %>%
  select(!bk_death_count)

# Manhattan
df_wave2_mn <- df %>%
  filter(date >= start_date & date <= end_date) %>%
  select(c(date, mn_death_count)) %>%
  mutate(
    t_i = get_diff(start_date, date),
    Y_i = cumsum(mn_death_count)
  ) %>%
  select(!mn_death_count)

# Queens
df_wave2_qn <- df %>%
  filter(date >= start_date & date <= end_date) %>%
  select(c(date, qn_death_count)) %>%
  mutate(
    t_i = get_diff(start_date, date),
    Y_i = cumsum(qn_death_count)
  ) %>%
  select(!qn_death_count)

# Staten Island
df_wave2_si <- df %>%
  filter(date >= start_date & date <= end_date) %>%
  select(c(date, si_death_count)) %>%
  mutate(
    t_i = get_diff(start_date, date),
    Y_i = cumsum(si_death_count)
  ) %>%
  select(!si_death_count)
```

### Check cumulative deaths over wave 2 for all of NYC

```{r}
df_wave2_nyc %>% 
  ggplot(aes(x = date, y = Y_i)) +
  geom_point() +
  scale_x_date(date_breaks = "1 month", date_labels = "%b")
```

### Run N-R

Running N-R for each borough and all of NYC using code from loaded R files

```{r}
# All of NYC
start_nyc2 <- get_start(df_wave2_nyc)
res_nyc2 <- NewtonRaphson(df_wave2_nyc, start_nyc2)
final_nyc2 <- res_nyc2[nrow(res_nyc2), 4:7]

# Bronx
start_bx2 <- get_start(df_wave2_bx)
res_bx2 <- NewtonRaphson(df_wave2_bx, start_bx2)
final_bx2 <- res_bx2[nrow(res_bx2), 4:7]

# Brooklyn
start_bk2 <- get_start(df_wave2_bk)
res_bk2 <- NewtonRaphson(df_wave2_bk, start_bk2)
final_bk2 <- res_bk2[nrow(res_bk2), 4:7]

# Manhattan
start_mn2 <- get_start(df_wave2_mn)
res_mn2 <- NewtonRaphson(df_wave2_mn, start_mn2)
final_mn2 <- res_mn2[nrow(res_mn2), 4:7]

# Queens
start_qn2 <- get_start(df_wave2_qn)
res_qn2 <- NewtonRaphson(df_wave2_qn, start_qn2)
final_qn2 <- res_qn2[nrow(res_qn2), 4:7]

# Staten Island
start_si2 <- get_start(df_wave2_si)
res_si2 <- NewtonRaphson(df_wave2_si, start_si2)
final_si2 <- res_si2[nrow(res_si2), 4:7]
```

### Plot final curves

```{r}
# Bronx
death_wave2_bx <- df_wave2_bx %>%
  mutate(
    Y_i_hat_start = N(t_i, start_bx2),
    Y_i_hat_final = N(t_i, final_bx2)
  ) %>%
  pivot_longer(Y_i_hat_start:Y_i_hat_final, names_to = "type", values_to = "value") %>%
  ggplot(aes(x = t_i)) + 
  geom_point(aes(y = Y_i), size = 1) +
  geom_line(aes(y = value, color = type)) +
  labs(title = "Bronx", x ="Days (since Sep 15, 2020)", y = "Cumulative deaths") +   
  scale_color_discrete(name = "Fitted model", labels = c("Final", "Start")) 

# Brooklyn
death_wave2_bk <- df_wave2_bk %>%
  mutate(
    Y_i_hat_start = N(t_i, start_bk2),
    Y_i_hat_final = N(t_i, final_bk2)
  ) %>%
  pivot_longer(Y_i_hat_start:Y_i_hat_final, names_to = "type", values_to = "value") %>%
  ggplot(aes(x = t_i)) + 
  geom_point(aes(y = Y_i), size = 1) +
  geom_line(aes(y = value, color = type)) +
  labs(title = "Brooklyn", x ="Days (since Sep 15, 2020)", y = "Cumulative deaths") +   
  scale_color_discrete(name = "Fitted model", labels = c("Final", "Start")) 

## Manhattan
death_wave2_mn <- df_wave2_mn %>%
  mutate(
    Y_i_hat_start = N(t_i, start_mn2),
    Y_i_hat_final = N(t_i, final_mn2)
  ) %>%
  pivot_longer(Y_i_hat_start:Y_i_hat_final, names_to = "type", values_to = "value") %>%
  ggplot(aes(x = t_i)) + 
  geom_point(aes(y = Y_i), size = 1) +
  geom_line(aes(y = value, color = type)) +
  labs(title = "Manhattan", x ="Days (since Sep 15, 2020)", y = "Cumulative deaths") +   
  scale_color_discrete(name = "Fitted model", labels = c("Final", "Start")) 

## Queens
death_wave2_qn <- df_wave2_qn %>%
  mutate(
    Y_i_hat_start = N(t_i, start_qn2),
    Y_i_hat_final = N(t_i, final_qn2)
  ) %>%
  pivot_longer(Y_i_hat_start:Y_i_hat_final, names_to = "type", values_to = "value") %>%
  ggplot(aes(x = t_i)) + 
  geom_point(aes(y = Y_i), size = 1) +
  geom_line(aes(y = value, color = type)) +
  labs(title = "Queens", x ="Days (since Sep 15, 2020)", y = "Cumulative deaths") +   
  scale_color_discrete(name = "Fitted model", labels = c("Final", "Start")) 

## Staten Island
death_wave2_si <- df_wave2_si %>%
  mutate(
    Y_i_hat_start = N(t_i, start_si2),
    Y_i_hat_final = N(t_i, final_si2)
  ) %>%
  pivot_longer(Y_i_hat_start:Y_i_hat_final, names_to = "type", values_to = "value") %>%
  ggplot(aes(x = t_i)) + 
  geom_point(aes(y = Y_i), size = 1) +
  geom_line(aes(y = value, color = type)) +
  labs(title = "Staten Island", x ="Days (since Sep 15, 2020)", y = "Cumulative deaths") +   
  scale_color_discrete(name = "Fitted model", labels = c("Final", "Start")) 

## NYC
death_wave2_nyc <- df_wave2_nyc %>%
  mutate(
    Y_i_hat_start = N(t_i, start_nyc2),
    Y_i_hat_final = N(t_i, final_nyc2)
  ) %>%
  pivot_longer(Y_i_hat_start:Y_i_hat_final, names_to = "type", values_to = "value") %>%
  ggplot(aes(x = t_i)) + 
  geom_point(aes(y = Y_i), size = 1) +
  geom_line(aes(y = value, color = type)) +
  labs(title = "New York City", x ="Days (since Sep 15, 2020)", 
       y = "Cumulative deaths") +   
  scale_color_discrete(name = "Fitted model", labels = c("Final", "Start")) 

ggarrange(death_wave2_nyc, death_wave2_bx, death_wave2_bk, death_wave2_mn, death_wave2_qn, death_wave2_si,  nrow = 2, ncol = 3, common.legend = TRUE, legend = "bottom") 
```

# Task 1.3

## NYC extended curve

```{r}
# Constructing a vector of future dates
more_days <- seq(from = as.Date("2020-12-12"), to = as.Date("2021-06-12"), 
                by = 'day')
more_days <- ymd(more_days)
currow <- nrow(df_wave2_nyc)
more_times <- seq(from = currow + 1, 
                  to = length(more_days) + currow)
Ys <- rep(NA, length(more_days))
more_data <- tibble(date = more_days, t_i = more_times, Y_i = Ys)

# Adding data to current data set
df_wave2_nyc_ext <- rbind(df_wave2_nyc, more_data)

death_wave2_nyc_ext <- df_wave2_nyc_ext %>%
  mutate(
    Y_i_hat_final = N(t_i, final_nyc2)
  ) %>%
  ggplot(aes(x = t_i)) + 
  geom_point(aes(y = Y_i), size = 1) +
  geom_line(aes(y = Y_i_hat_final, color = "red")) +
  labs(title = "New York City", x ="Days (since Sep 15, 2020)", y = "Cumulative deaths") +
  theme(legend.position = "none") +
  coord_cartesian(xlim = c(0, 300))
```

## BX extended curve

```{r}
# Constructing a vector of future dates
more_days <- seq(from = as.Date("2020-12-12"), to = as.Date("2021-06-12"), 
                by = 'day')
more_days <- ymd(more_days)
currow <- nrow(df_wave2_bx)
more_times <- seq(from = currow + 1, 
                  to = length(more_days) + currow)
Ys <- rep(NA, length(more_days))
more_data <- tibble(date = more_days, t_i = more_times, Y_i = Ys)

# Adding data to current data set
df_wave2_bx_ext <- rbind(df_wave2_bx, more_data)

# Saving population value
pop_bx <- 1472654

# Making new plot
death_wave2_bx_ext <- df_wave2_bx_ext %>%
  mutate(
    Y_i_hat_final = N(t_i, final_bx2)
  ) %>%
  ggplot(aes(x = t_i)) + 
  geom_point(aes(y = (Y_i/pop_bx)*100000), size = 1) +
  geom_line(aes(y = (Y_i_hat_final/pop_bx)*100000), color = "red") +
  labs(title = "Bronx", x ="Days (since Sep 15, 2020)", 
       y = "Cumulative deaths per 100,000 capita") +
  theme(legend.position = "none") +
  coord_cartesian(xlim = c(0, 300), ylim = c(0, 45))
```

## BK extended curve

```{r}
# Constructing a vector of future dates
more_days <- seq(from = as.Date("2020-12-12"), to = as.Date("2021-06-12"), 
                by = 'day')
more_days <- ymd(more_days)
currow <- nrow(df_wave2_bk)
more_times <- seq(from = currow + 1, 
                  to = length(more_days) + currow)
Ys <- rep(NA, length(more_days))
more_data <- tibble(date = more_days, t_i = more_times, Y_i = Ys)

# Adding data to current data set
df_wave2_bk_ext <- rbind(df_wave2_bk, more_data)

# Saving population value
pop_bk <- 2736074

# Making new plot
death_wave2_bk_ext <- df_wave2_bk_ext %>%
  mutate(
    Y_i_hat_final = N(t_i, final_bk2)
  ) %>%
  ggplot(aes(x = t_i)) + 
  geom_point(aes(y = (Y_i/pop_bk)*100000), size = 1) +
  geom_line(aes(y = (Y_i_hat_final/pop_bk)*100000), color = "red") +
  labs(title = "Brooklyn", x ="Days (since Sep 15, 2020)", 
       y = "Cumulative deaths per 100,000 capita") +
  theme(legend.position = "none") +
  coord_cartesian(xlim = c(0, 300), ylim = c(0, 45))
```

## MN extended curve

```{r}
# Constructing a vector of future dates
more_days <- seq(from = as.Date("2020-12-12"), to = as.Date("2021-06-12"), 
                by = 'day')
more_days <- ymd(more_days)
currow <- nrow(df_wave2_mn)
more_times <- seq(from = currow + 1, 
                  to = length(more_days) + currow)
Ys <- rep(NA, length(more_days))
more_data <- tibble(date = more_days, t_i = more_times, Y_i = Ys)

# Adding data to current data set
df_wave2_mn_ext <- rbind(df_wave2_mn, more_data)

# Saving population value
pop_mn <- 1694251

# Making new plot
death_wave2_mn_ext <- df_wave2_mn_ext %>%
  mutate(
    Y_i_hat_final = N(t_i, final_mn2)
  ) %>%
  ggplot(aes(x = t_i)) + 
  geom_point(aes(y = (Y_i/pop_mn)*100000), size = 1) +
  geom_line(aes(y = (Y_i_hat_final/pop_mn)*100000), color = "red") +
  labs(title = "Manhattan", x ="Days (since Sep 15, 2020)", 
       y = "Cumulative deaths per 100,000 capita") +
  theme(legend.position = "none") +
  coord_cartesian(xlim = c(0, 300), ylim = c(0, 45))
```

## QN extended curve

```{r}
# Constructing a vector of future dates
more_days <- seq(from = as.Date("2020-12-12"), to = as.Date("2021-06-12"), 
                by = 'day')
more_days <- ymd(more_days)
currow <- nrow(df_wave2_qn)
more_times <- seq(from = currow + 1, 
                  to = length(more_days) + currow)
Ys <- rep(NA, length(more_days))
more_data <- tibble(date = more_days, t_i = more_times, Y_i = Ys)

# Adding data to current data set
df_wave2_qn_ext <- rbind(df_wave2_qn, more_data)

# Saving population value
pop_qn <- 2405464

# Making new plot
death_wave2_qn_ext <- df_wave2_qn_ext %>%
  mutate(
    Y_i_hat_final = N(t_i, final_qn2)
  ) %>%
  ggplot(aes(x = t_i)) + 
  geom_point(aes(y = (Y_i/pop_qn)*100000), size = 1) +
  geom_line(aes(y = (Y_i_hat_final/pop_qn)*100000), color = "red") +
  labs(title = "Queens", x ="Days (since Sep 15, 2020)", 
       y = "Cumulative deaths per 100,000 capita") +
  theme(legend.position = "none") +
  coord_cartesian(xlim = c(0, 300), ylim = c(0, 45))
```

## SI extended curve

```{r}
# Constructing a vector of future dates
more_days <- seq(from = as.Date("2020-12-12"), to = as.Date("2021-06-12"), 
                by = 'day')
more_days <- ymd(more_days)
currow <- nrow(df_wave2_si)
more_times <- seq(from = currow + 1, 
                  to = length(more_days) + currow)
Ys <- rep(NA, length(more_days))
more_data <- tibble(date = more_days, t_i = more_times, Y_i = Ys)

# Adding data to current data set
df_wave2_si_ext <- rbind(df_wave2_si, more_data)

# Saving population value
pop_si <- 495747

# Making new plot
death_wave2_si_ext <- df_wave2_si_ext %>%
  mutate(
    Y_i_hat_final = N(t_i, final_si2)
  ) %>%
  ggplot(aes(x = t_i)) + 
  geom_point(aes(y = (Y_i/pop_si)*100000), size = 1) +
  geom_line(aes(y = (Y_i_hat_final/pop_si)*100000), color = "red") +
  labs(title = "Staten Island", x ="Days (since Sep 15, 2020)", 
       y = "Cumulative deaths per 100,000 capita") +
  theme(legend.position = "none") +
  coord_cartesian(xlim = c(0, 300), ylim = c(0, 45))
```

## Showing all extended curves

```{r}
ggarrange(death_wave2_bx_ext, 
          death_wave2_bk_ext, death_wave2_mn_ext, 
          death_wave2_qn_ext, death_wave2_si_ext,  
          nrow = 2, ncol = 3, common.legend = TRUE, legend = "bottom") 
```

